---
title: "Lesson 5: Small-Population Confidence Intervals for Differences in Means"
subtitle: "Statistical Estimation for Data Science and AI"
---


::: {.column-margin}
{{< video 01_small-sample-confidence-intervals-for-the-difference-between-population-means.mp4 >}}

[Transcript](01_small-sample-confidence-intervals-for-the-difference-between-population-means.en.en.txt)
[Handout](02_video-slides-for-small-sample-confidence-intervals-for-the-difference-between_module4_lesson5.pdf){width=700 height=980}
:::

## Slides on Small-Sample CI for the Difference Between Population Means {#sec-small-sample-ci-difference-means}

### Setting and Assumptions {#sec-small-sample-ci-setting}

Suppose that
$X_{1,1},X_{1,2},\ldots,X_{1,n_1} \overset{\mathrm{iid}}{\sim} \mathcal{N}(\mu_1,\sigma_1^2)$
and
$X_{2,1},X_{2,2},\ldots,X_{2,n_2} \overset{\mathrm{iid}}{\sim} \mathcal{N}(\mu_2,\sigma_2^2)$
The two samples are independent, both population variances are unknown, and one or both sample sizes are small:
$n_1 \leq 30 \qquad \text{and/or} \qquad n_2 \leq 30$
The goal is to construct a $100(1-\alpha)\%$ confidence interval for
$\mu_1-\mu_2$

::: {.callout-important}
## Equal-Variance Assumption

The pooled two-sample procedure requires the strong assumption

$$
\color{red}{\sigma_1^2=\sigma_2^2=\sigma^2}
$$

where $\sigma^2$ denotes the unknown common population variance
:::

## The Estimator {#sec-small-sample-ci-estimator}

The natural estimator of the difference between the population means is

$$
\widehat{\mu_1-\mu_2}=\overline{X}_1-\overline{X}_2
$$

The two sample means have distributions

$$
\overline{X}_1
\sim
\mathcal{N}\!\left(\mu_1,\frac{\sigma_1^2}{n_1}\right)
$$

and

$$
\overline{X}_2
\sim
\mathcal{N}\!\left(\mu_2,\frac{\sigma_2^2}{n_2}\right)
$$

Because the samples are independent, $\overline{X}_1$ and $\overline{X}_2$ are independent

### Mean of the Estimator {#sec-small-sample-ci-estimator-mean}

$$
\begin{aligned}
\mathbb{E}\!\left[\overline{X}_1-\overline{X}_2\right]
&=
\mathbb{E}\!\left[\overline{X}_1\right]
-
\mathbb{E}\!\left[\overline{X}_2\right] \\
&= \mu_1-\mu_2
\end{aligned}
$$

Thus, $\overline{X}_1-\overline{X}_2$ is an unbiased estimator of $\mu_1-\mu_2$

### Variance of the Estimator {#sec-small-sample-ci-estimator-variance}

$$
\begin{aligned}
\operatorname{Var}\!\left(\overline{X}_1-\overline{X}_2\right)
&=
\operatorname{Var}\!\left(\overline{X}_1+(-1)\overline{X}_2\right) \\
&=
\operatorname{Var}\!\left(\overline{X}_1\right)
+
\operatorname{Var}\!\left((-1)\overline{X}_2\right)
&& \text{independence} \\
&=
\operatorname{Var}\!\left(\overline{X}_1\right)
+(-1)^2\operatorname{Var}\!\left(\overline{X}_2\right) \\
&=
\frac{\sigma_1^2}{n_1}
+
\frac{\sigma_2^2}{n_2}
\end{aligned}
$$

Therefore,

$$
\overline{X}_1-\overline{X}_2
\sim
\mathcal{N}\!\left(
\mu_1-\mu_2,
\frac{\sigma_1^2}{n_1}+\frac{\sigma_2^2}{n_2}
\right)
$$

Under the equal-variance assumption,

$$
\frac{\sigma_1^2}{n_1}+\frac{\sigma_2^2}{n_2}
=
\sigma^2\!\left(\frac{1}{n_1}+\frac{1}{n_2}\right)
$$

Hence the standardized statistic is

$$
Z
=
\frac{
\overline{X}_1-\overline{X}_2-(\mu_1-\mu_2)
}{
\sigma\sqrt{\frac{1}{n_1}+\frac{1}{n_2}}
}
\sim
\mathcal{N}(0,1)
$$

The difficulty is that the common variance $\sigma^2$ is unknown

## The Pooled Variance {#sec-small-sample-ci-pooled-variance}

The sample variances $S_1^2$ and $S_2^2$ are independent estimators of the common variance $\sigma^2$. They should not be combined using an unweighted average because the estimates may be based on different sample sizes

::: {#def-pooled-variance}
## Pooled Variance

The pooled estimator of the common variance is the degrees-of-freedom-weighted average

$$
S_p^2
=
\frac{
\underbrace{(n_1-1)S_1^2}_{\text{sample 1 sum of squares}}
+
\underbrace{(n_2-1)S_2^2}_{\text{sample 2 sum of squares}}
}{
\underbrace{n_1+n_2-2}_{\text{total degrees of freedom}}
}
$$
:::

For independent normal samples,

$$
\frac{(n_1-1)S_1^2}{\sigma^2}
\sim
\chi^2_{n_1-1}
$$

and

$$
\frac{(n_2-1)S_2^2}{\sigma^2}
\sim
\chi^2_{n_2-1}
$$

These two chi-squared random variables are independent. Therefore,

$$
\begin{aligned}
\frac{(n_1-1)S_1^2}{\sigma^2}
+
\frac{(n_2-1)S_2^2}{\sigma^2}
&\sim
\chi^2_{(n_1-1)+(n_2-1)} \\
&=
\chi^2_{n_1+n_2-2}
\end{aligned}
$$

From the definition of $S_p^2$,

$$
\begin{aligned}
\frac{(n_1+n_2-2)S_p^2}{\sigma^2}
&=
\frac{(n_1-1)S_1^2+(n_2-1)S_2^2}{\sigma^2} \\
&=
\frac{(n_1-1)S_1^2}{\sigma^2}
+
\frac{(n_2-1)S_2^2}{\sigma^2} \\
&\sim
\chi^2_{n_1+n_2-2}
\end{aligned}
$$

## Constructing the Pivotal Quantity {#sec-small-sample-ci-pivotal-quantity}

Let

$$
\nu=n_1+n_2-2
$$

The standardized normal statistic can be divided by the square root of an independent chi-squared statistic divided by its degrees of freedom:

$$
\begin{aligned}
T
&=
\frac{
\overline{X}_1-\overline{X}_2-(\mu_1-\mu_2)
}{
S_p\sqrt{\frac{1}{n_1}+\frac{1}{n_2}}
} \\
&=
\frac{
\dfrac{
\overline{X}_1-\overline{X}_2-(\mu_1-\mu_2)
}{
\sigma\sqrt{\frac{1}{n_1}+\frac{1}{n_2}}
}
}{
\sqrt{\dfrac{S_p^2}{\sigma^2}}
} \\
&=
\frac{
\underbrace{
\dfrac{
\overline{X}_1-\overline{X}_2-(\mu_1-\mu_2)
}{
\sigma\sqrt{\frac{1}{n_1}+\frac{1}{n_2}}
}
}_{\displaystyle Z\sim\mathcal{N}(0,1)}
}{
\underbrace{
\sqrt{
\dfrac{
\left((n_1+n_2-2)S_p^2\right)/\sigma^2
}{
 n_1+n_2-2
}
}
}_{\displaystyle \sqrt{\chi_\nu^2/\nu}}
}
\end{aligned}
$$

For normal samples, the sample means are independent of the sample variances. Consequently, the numerator and denominator above are independent, and

$$
T
=
\frac{
\overline{X}_1-\overline{X}_2-(\mu_1-\mu_2)
}{
\sqrt{S_p^2\left(\frac{1}{n_1}+\frac{1}{n_2}\right)}
}
\sim
t_{n_1+n_2-2}
$$

## Equal-Variance Confidence Interval {#sec-small-sample-ci-equal-variance-interval}

Let $t_{\alpha/2,\nu}$ denote the upper $\alpha/2$ critical value of a Student's $t$ distribution with $\nu$ degrees of freedom. Then

$$
\mathbb{P}\!\left(
-t_{\alpha/2,\nu}
<
T
<
t_{\alpha/2,\nu}
\right)
=1-\alpha
$$

Substituting the pivotal quantity gives

$$
-t_{\alpha/2,n_1+n_2-2}
<
\frac{
\overline{X}_1-\overline{X}_2-(\mu_1-\mu_2)
}{
S_p\sqrt{\frac{1}{n_1}+\frac{1}{n_2}}
}
<
t_{\alpha/2,n_1+n_2-2}
$$

::: {#thm-small-sample-ci-equal-variance}
## Small-Sample Confidence Interval for $\mu_1-\mu_2$

For two independent normal samples with unknown but equal population variances, a $100(1-\alpha)\%$ confidence interval for $\mu_1-\mu_2$ is

$$
\boxed{
(\overline{X}_1-\overline{X}_2)
\pm
t_{\alpha/2,n_1+n_2-2}
S_p\sqrt{\frac{1}{n_1}+\frac{1}{n_2}}
}
$$

where

$$
S_p^2
=
\frac{(n_1-1)S_1^2+(n_2-1)S_2^2}{n_1+n_2-2}
$$
:::

## Example: A 90% Confidence Interval {#sec-small-sample-ci-example}

::: {#exm-small-sample-ci-difference}
## Confidence Interval for $\mu_1-\mu_2$

Suppose the two samples have summaries

$$
\begin{aligned}
n_1 &= 9
&\qquad \overline{x}_1 &= 23.2
&\qquad s_1^2 &= 4.3 \\
n_2 &= 8
&\qquad \overline{x}_2 &= 24.7
&\qquad s_2^2 &= 5.2
\end{aligned}
$$

For a $90\%$ confidence interval,

$$
\alpha=0.10
\qquad
\frac{\alpha}{2}=0.05
\qquad
\nu=n_1+n_2-2=15
$$

First compute the pooled variance:

$$
\begin{aligned}
s_p^2
&=
\frac{(n_1-1)s_1^2+(n_2-1)s_2^2}{n_1+n_2-2} \\
&=
\frac{(9-1)(4.3)+(8-1)(5.2)}{9+8-2} \\
&=
\frac{34.4+36.4}{15} \\
&=
4.72
\end{aligned}
$$

The critical value is

$$
t_{0.05,15}=1.753
$$

The confidence interval is

$$
\begin{aligned}
(\overline{x}_1-\overline{x}_2)
&\pm
t_{0.05,15}
\sqrt{s_p^2\left(\frac{1}{n_1}+\frac{1}{n_2}\right)} \\
&=
23.2-24.7
\pm
1.753\sqrt{4.72\left(\frac{1}{9}+\frac{1}{8}\right)} \\
&=
-1.5\pm1.851 \\
&=
\boxed{(-3.351,\ 0.351)}
\end{aligned}
$$

:::

The critical value can be obtained in R with

```{r}
qt(0.95, df = 15)
```

## Unequal Population Variances {#sec-small-sample-ci-unequal-variances}

When $\sigma_1^2$ and $\sigma_2^2$ cannot reasonably be assumed equal, there is no exact confidence interval of the same simple form. This is associated with the Behrens--Fisher problem

### Welch's Approximation {#sec-small-sample-ci-welch}

Welch's method replaces the pooled standard error with

$$
\sqrt{\frac{S_1^2}{n_1}+\frac{S_2^2}{n_2}}
$$

and uses the approximate pivotal quantity

$$
T_W
=
\frac{
\overline{X}_1-\overline{X}_2-(\mu_1-\mu_2)
}{
\sqrt{\frac{S_1^2}{n_1}+\frac{S_2^2}{n_2}}
}
\overset{\mathrm{approx}}{\sim}
t_\nu
$$

The Welch--Satterthwaite degrees of freedom are

$$
\nu
=
\frac{
\left(\dfrac{S_1^2}{n_1}+\dfrac{S_2^2}{n_2}\right)^2
}{
\dfrac{\left(S_1^2/n_1\right)^2}{n_1-1}
+
\dfrac{\left(S_2^2/n_2\right)^2}{n_2-1}
}
$$

::: {#thm-small-sample-ci-welch}
## Welch Confidence Interval

An approximate $100(1-\alpha)\%$ confidence interval for $\mu_1-\mu_2$ is

$$
\boxed{
(\overline{X}_1-\overline{X}_2)
\pm
t_{\alpha/2,\nu}
\sqrt{\frac{S_1^2}{n_1}+\frac{S_2^2}{n_2}}
}
$$

where $\nu$ is given by the Welch--Satterthwaite approximation
:::

R uses Welch's procedure by default for a two-sample $t$ interval:

```{r}
x <- rnorm(10)
y <- rnorm(14)

t.test(x, y, conf.level = 0.90)
```

One run shown in the slides produced

```text
Welch Two Sample t-test

data:  x and y
t = 0.030583, df = 16.916, p-value = 0.976
alternative hypothesis: true difference in means is not equal to 0
90 percent confidence interval:
 -0.6289463  0.6514495
sample estimates:
 mean of x   mean of y
-0.2765715  -0.2878232
```


## Reflection

