---
title: "Lesson 3: A Confidence Interval for a Ratio of Variances"
subtitle: "Statistical Estimation for Data Science and AI"
---


::: {.column-margin}
{{< video 01_a-confidence-interval-for-a-ratio-of-variances.mp4 >}}

[Transcript](01_a-confidence-interval-for-a-ratio-of-variances.en.txt)
[Handout](02_video-slides-for-a-confidence-interval-for-a-ratio-of-variances_module5_lesson3.pdf){width=700 height=980}
:::

## A CI for a Ratio of Variances {#sec-ci-ratio-variances}

### The F Distribution {#sec-f-distribution}

::: {#def-f-distribution}
## F Distribution

Suppose $X_1$ and $X_2$ are independent random variables with

$$
X_1\sim\chi^2_{n_1}
\qquad
X_2\sim\chi^2_{n_2}
$$

Define

$$
F
=
\frac{X_1/n_1}{X_2/n_2}
$$

Then

$$
F\sim F(n_1,n_2)
$$
:::

The $F$ distribution is positive and generally right-skewed. With small denominator degrees of freedom, it has a particularly heavy right tail. As both degrees of freedom increase, it becomes increasingly concentrated around 1.

The four shapes shown in the slides can be reproduced with

```{r}
#| fig-width: 8
#| fig-height: 7

parameters <- list(
  c(df1 = 1,  df2 = 1,  xmax = 20),
  c(df1 = 5,  df2 = 5,  xmax = 20),
  c(df1 = 50, df2 = 50, xmax = 5),
  c(df1 = 10, df2 = 2,  xmax = 5)
)

old_par <- par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

for (pars in parameters) {
  curve(
    df(x, df1 = pars["df1"], df2 = pars["df2"]),
    from = 0,
    to = pars["xmax"],
    n = 1000,
    xlab = "f",
    ylab = "Density",
    main = sprintf("df1 = %g, df2 = %g", pars["df1"], pars["df2"])
  )
}

par(old_par)
```

For $x>0$, its probability density function is

$$
f(x;n_1,n_2)
=
\frac{1}{B(n_1/2,n_2/2)}
\left(\frac{n_1}{n_2}\right)^{n_1/2}
 x^{n_1/2-1}
\left(1+\frac{n_1}{n_2}x\right)^{-(n_1+n_2)/2}
$$

Its mean is

$$
\mathbb{E}[F]
=
\frac{n_2}{n_2-2}
\qquad
n_2>2
$$

and its variance is

$$
\operatorname{Var}(F)
=
\frac{2n_2^2(n_1+n_2-2)}
{n_1(n_2-2)^2(n_2-4)}
\qquad
n_2>4
$$

In R, the 95th percentile and the corresponding cumulative probability for an $F(5,1)$ random variable are

```{r}
qf(0.95, df1 = 5, df2 = 1)
pf(6.608, df1 = 5, df2 = 1)
```

which give approximately

$$
q_F(0.95)=6.608
\qquad
P(F\leq 6.608)=0.9499824
$$

### Deriving the Mean of the F Distribution {#sec-mean-f-distribution}

Because $X_1$ and $X_2$ are independent,

$$
\begin{aligned}
\mathbb{E}[F]
&=
\mathbb{E}\!\left[
\frac{X_1/n_1}{X_2/n_2}
\right] \\
&=
\frac{n_2}{n_1}
\mathbb{E}\!\left[
\frac{X_1}{X_2}
\right] \\
&=
\frac{n_2}{n_1}
\mathbb{E}[X_1]
\mathbb{E}\!\left[\frac{1}{X_2}\right]
&& \text{independence} \\
&=
n_2
\mathbb{E}\!\left[\frac{1}{X_2}\right]
&& \mathbb{E}[X_1]=n_1
\end{aligned}
$$

Since $X_2\sim\chi^2_{n_2}$,

$$
\begin{aligned}
\mathbb{E}\!\left[\frac{1}{X_2}\right]
&=
\int_0^\infty
\frac{1}{x}
\frac{1}{2^{n_2/2}\Gamma(n_2/2)}
 x^{n_2/2-1}e^{-x/2}
\,dx \\
&=
\frac{1}{2^{n_2/2}\Gamma(n_2/2)}
\int_0^\infty
x^{n_2/2-2}e^{-x/2}
\,dx \\
&=
\frac{2^{n_2/2-1}\Gamma(n_2/2-1)}
{2^{n_2/2}\Gamma(n_2/2)} \\
&=
\frac{\Gamma(n_2/2-1)}
{2\Gamma(n_2/2)} \\
&=
\frac{1}{n_2-2}
&& \Gamma(a)=(a-1)\Gamma(a-1)
\end{aligned}
$$

Therefore,

$$
\boxed{
\mathbb{E}[F]
=
\frac{n_2}{n_2-2}
}
$$

### Ratio of Two Population Variances {#sec-ratio-population-variances}

Suppose

$$
X_{11},X_{12},\ldots,X_{1n_1}
\overset{\mathrm{iid}}{\sim}
\mathcal{N}(\mu_1,\sigma_1^2)
$$

and independently

$$
X_{21},X_{22},\ldots,X_{2n_2}
\overset{\mathrm{iid}}{\sim}
\mathcal{N}(\mu_2,\sigma_2^2)
$$

Let $S_1^2$ and $S_2^2$ be the two sample variances. Then

$$
\frac{(n_1-1)S_1^2}{\sigma_1^2}
\sim
\chi^2_{n_1-1}
$$

and

$$
\frac{(n_2-1)S_2^2}{\sigma_2^2}
\sim
\chi^2_{n_2-1}
$$

These two random variables are independent. Hence

$$
\begin{aligned}
F
&:=
\frac{
\left[(n_1-1)S_1^2/\sigma_1^2\right]/(n_1-1)
}{
\left[(n_2-1)S_2^2/\sigma_2^2\right]/(n_2-1)
} \\
&=
\frac{\sigma_2^2}{\sigma_1^2}
\frac{S_1^2}{S_2^2}
\end{aligned}
$$

with

$$
F\sim F(n_1-1,n_2-1)
$$

::: {.callout-note}
## F Critical-Value Notation

Here, $F_{\gamma,\nu_1,\nu_2}$ denotes an upper-tail critical value satisfying

$$
P\!\left(
F_{\nu_1,\nu_2}>F_{\gamma,\nu_1,\nu_2}
\right)
=
\gamma
$$
:::

A central $100(1-\alpha)\%$ probability statement is

$$
F_{1-\alpha/2,n_1-1,n_2-1}
<
\frac{\sigma_2^2}{\sigma_1^2}
\frac{S_1^2}{S_2^2}
<
F_{\alpha/2,n_1-1,n_2-1}
$$

Solving for $\sigma_1^2/\sigma_2^2$ gives

::: {#def-ci-ratio-variances}
## Confidence Interval for a Ratio of Variances

A $100(1-\alpha)\%$ confidence interval for $\sigma_1^2/\sigma_2^2$ is

$$
\boxed{
\left(
\frac{S_1^2/S_2^2}
{F_{\alpha/2,n_1-1,n_2-1}},
\frac{S_1^2/S_2^2}
{F_{1-\alpha/2,n_1-1,n_2-1}}
\right)
}
$$
:::

::: {#exm-county-score-variances}
## County Placement Scores

Fifth-grade students from two neighboring counties took a placement exam.

Group 1, from County A, contained 18 students and had a sample mean score of 77.2. Group 2, from County B, contained 15 students and had a sample mean score of 75.3.

The reported variance estimates are

$$
s_1^2=15.3
\qquad
s_2^2=19.7
$$

A pooled-variance two-sample $t$ procedure would require $\sigma_1^2=\sigma_2^2$. Construct a 99% confidence interval for $\sigma_1^2/\sigma_2^2$ and assess whether equality is plausible.
:::

The sample information is

$$
n_1=18
\qquad
n_2=15
\qquad
\nu_1=17
\qquad
\nu_2=14
$$

The ratio of sample variances is

$$
\frac{s_1^2}{s_2^2}
=
\frac{15.3}{19.7}
\approx
0.77665
$$

::: {.callout-warning}
## Numerical inconsistency in the slides

The slides print

$$
F_{0.005,17,14}=3.98267
\qquad
F_{0.995,17,14}=1.00217
$$

and consequently report

$$
\boxed{(0.19501,0.77497)}
$$

These are not the two-sided 99% critical values for an $F(17,14)$ distribution.
:::

Using the stated degrees of freedom, the correct upper-tail critical values are

$$
F_{0.005,17,14}=4.15916
\qquad
F_{0.995,17,14}=0.26011
$$

They can be obtained in R by

```{r}
qf(c(0.995, 0.005), df1 = 17, df2 = 14)
```

The corrected interval is

$$
\begin{aligned}
\frac{15.3/19.7}{4.15916}
&<
\frac{\sigma_1^2}{\sigma_2^2}
<
\frac{15.3/19.7}{0.26011} \\
0.18673
&<
\frac{\sigma_1^2}{\sigma_2^2}
<
2.98582
\end{aligned}
$$

Thus,

$$
\boxed{(0.18673,2.98582)}
$$

Because this corrected interval contains 1, equality of the population variances remains plausible at the 99% confidence level.

## Reflection

