---
title: "Lesson 1: Introduction to Confidence Intervals"
subtitle: "Statistical Estimation for Data Science and AI"
---


::: {.column-margin}
{{< video 01_lets-build-a-confidence-interval.mp4 >}}

[Transcript](01_lets-build-a-confidence-interval.en.txt)
[Handout](02_video-slides-for-lets-build-a-confidence-interval_module4_lesson1.pdf){width=700 height=980}
:::


## Slides on Introduction to Confidence Intervals {#sec-introduction-confidence-intervals}

Suppose a reported interval is

> A 95% confidence interval for the mean $\mu$ is $(-2.14, 3.07)$

::: {.callout-warning}
## What this does not mean

This statement does **not** mean that

1. we are 95% confident that the fixed parameter $\mu$ lies between $-2.14$ and $3.07$
2. the probability that $\mu$ lies between $-2.14$ and $3.07$ is $0.95$

Once the sample has been observed, both endpoints are fixed and the parameter $\mu$ is fixed
:::

The randomness lies in the sampling procedure:

1. collect a random sample
2. estimate the parameter
3. construct a confidence interval

Repeating this procedure with a new random sample generally produces a different interval.

::: {#def-confidence-interval-coverage}
## Confidence-Interval Coverage

Repeated samples produce repeated confidence intervals. A 95% confidence-interval procedure is constructed so that, over repeated sampling, 95% of the resulting intervals contain the true parameter $\mu$
:::

## Building a Confidence Interval from the Ground Up {#sec-building-confidence-interval}

Suppose that $X_1, X_2, \ldots, X_n \overset{\mathrm{iid}}{\sim} \mathcal{N}(\mu,\sigma^2)$ and assume that the population variance $\sigma^2$ is known.

The sample mean $\overline{X}$ is an estimator of $\mu$, and

$$
\overline{X} \sim \mathcal{N}\!\left(\mu,\frac{\sigma^2}{n}\right)
$$

Therefore, the standardized sample mean satisfies

$$
Z = \frac{\overline{X}-\mu}{\sigma/\sqrt{n}} \sim \mathcal{N}(0,1)
$$

## The Central 95% of the Standard Normal Distribution {#sec-central-95-standard-normal}

For $Z\sim\mathcal{N}(0,1)$, the central 95% of the distribution lies between $-1.96$ and $1.96$:

$$
P(-1.96 < Z < 1.96)=0.95
$$

The upper endpoint is the $0.975$ quantile because the remaining probability $0.05$ is divided equally between the two tails:

```{r}
qnorm(0.975)
```

which gives approximately

$$
1.96
$$

## Deriving the 95% Confidence Interval {#sec-deriving-95-confidence-interval}

Substitute the standardized sample mean for $Z$ and solve the probability statement for $\mu$:

$$
\begin{aligned}
P(-1.96 < Z < 1.96)
&= 0.95 \\
P\!\left(
-1.96
<
\frac{\overline{X}-\mu}{\sigma/\sqrt{n}}
<
1.96
\right)
&= 0.95
&& \text{substitute the standardized sample mean} \\
P\!\left(
-1.96\frac{\sigma}{\sqrt{n}}
<
\overline{X}-\mu
<
1.96\frac{\sigma}{\sqrt{n}}
\right)
&= 0.95
&& \text{multiply by }\frac{\sigma}{\sqrt n} \\
P\!\left(
\overline{X}-1.96\frac{\sigma}{\sqrt{n}}
<
\mu
<
\overline{X}+1.96\frac{\sigma}{\sqrt{n}}
\right)
&= 0.95
&& \text{solve the inequalities for }\mu
\end{aligned}
$$

Thus, a 95% confidence interval for the mean $\mu$ is

$$
\left(
\overline{X}-1.96\frac{\sigma}{\sqrt{n}},
\overline{X}+1.96\frac{\sigma}{\sqrt{n}}
\right)
$$

The same interval can be written as

$$
\overline{X} \pm \underbrace{1.96\frac{\sigma}{\sqrt{n}}}_{\text{margin of error}}
$$

## Critical Values {#sec-critical-values}

::: {#def-critical-value}
## Critical Value

A critical value is a value that cuts off a specified area under a probability density function
:::

Let $Z\sim\mathcal{N}(0,1)$. The critical value $z_\alpha$ is defined so that an area $\alpha$ lies to its right:

$$
P(Z>z_\alpha)=\alpha
$$

Equivalently,

$$
P(Z\leq z_\alpha)=1-\alpha
$$

For a two-sided confidence interval with confidence coefficient $1-\alpha$, each tail contains probability $\alpha/2$:

$$
P\!\left(-z_{\alpha/2}< Z <z_{\alpha/2}\right)=1-\alpha
$$

By symmetry of the standard normal distribution,

$$
-z_{\alpha/2}=z_{1-\alpha/2}
$$

For a 95% confidence interval,

$$
\begin{aligned}
1-\alpha &= 0.95 \\
\alpha &= 0.05 \\
\frac{\alpha}{2} &= 0.025 \\
z_{\alpha/2} &= z_{0.025}=1.96
\end{aligned}
$$

Here, the notation $z_{0.025}$ denotes the positive critical value that leaves area $0.025$ to its right.

## Confidence Interval for a Normal Mean with Known Variance {#sec-normal-mean-known-variance}

::: {#thm-normal-mean-confidence-interval}
## Normal Mean with Known Variance

Suppose that $X_1, X_2, \ldots, X_n \overset{\mathrm{iid}}{\sim} \mathcal{N}(\mu,\sigma^2)$ where $\sigma^2$ is known. 
A $100(1-\alpha)\%$ confidence interval for $\mu$ is

$$
\boxed{ \overline{X} \pm \underbrace{z_{\alpha/2}\frac{\sigma}{\sqrt{n}}}_{\text{margin of error} } }
$$

Equivalently, the interval is

$$
\left(
\overline{X}-z_{\alpha/2}\frac{\sigma}{\sqrt{n}},
\overline{X}+z_{\alpha/2}\frac{\sigma}{\sqrt{n}}
\right)
$$
:::

## Extension Using the Central Limit Theorem {#sec-confidence-interval-clt}

The exact derivation above depends on the sample coming from a normal distribution, because then $\overline{X}$ is exactly normally distributed.

For a more general population distribution, the Central Limit Theorem implies that $\overline{X}$ is approximately normally distributed when the sample size is large. The slides use $n>30$ as a rule of thumb.

::: {#thm-large-sample-confidence-interval}
## Large-Sample Confidence Interval

Suppose that

$$
X_1, X_2, \ldots, X_n
$$

is a random sample from any distribution with mean $\mu$ and finite variance

$$
\sigma^2<\infty
$$

For large $n$, an approximate $100(1-\alpha)\%$ confidence interval for $\mu$ is

$$
\boxed{
\overline{X}
\pm
z_{\alpha/2}\frac{\sigma}{\sqrt{n}}
}
$$
:::

## Example: Monthly Spending on Organic Produce {#sec-example-organic-produce}

::: {#exm-organic-produce-confidence-interval}
## Organic-Produce Spending

A supermarket chain is considering adding more organic produce in a particular region. An external marketing company obtains a random sample of $200$ customers from the region.

The observed average monthly spending on organic produce is $36$ dollars per person. Based on previous studies, the population variance of spending is taken to be $5$.

Find a 90% confidence interval for the true mean monthly spending on organic produce among all customers in the region.

The supplied values are

$$
\begin{aligned}
n &= 200 \\
\overline{x} &= 36 \\
\sigma^2 &= 5 \\
\sigma &= \sqrt{5}
\end{aligned}
$$

For a 90% confidence interval,

$$
\begin{aligned}
1-\alpha &= 0.90 \\
\alpha &= 0.10 \\
\frac{\alpha}{2} &= 0.05
\end{aligned}
$$

The required critical value is

```{r}
qnorm(0.95)
```

which gives

$$
z_{\alpha/2}=z_{0.05}=1.645
$$

Substitute the values into the confidence-interval formula:

$$
\begin{aligned}
\overline{x}
\pm
z_{\alpha/2}\frac{\sigma}{\sqrt{n}}
&=
36
\pm
1.645\frac{\sqrt{5}}{\sqrt{200}} \\
&=
36
\pm
\underbrace{0.260}_{\text{margin of error}} \\
&=
(35.74,36.26)
\end{aligned}
$$

Thus, the 90% confidence interval for the population mean monthly spending is

$$
\boxed{(35.74,36.26)}
$$
:::


## Reflection

