---
title: "Lesson 4:  The Central Limit Theorem"
subtitle: "Statistical Estimation for Data Science and AI"
---

::: {.column-margin}
{{< video 01_the-central-limit-theorem.mp4 >}}

[Transcript](01_the-central-limit-theorem.en.txt)
[Handout](02_video-slides-for-the-central-limit-theorem_module_3_lesson4.pdf){width=700 height=980}
:::

## Slides on CLT

###Linear Combinations of Normal Random Variables {#sec-linear-combinations-normal}

::: {#thm-linear-combinations-normal}
## Linear Combinations of Normal Random Variables

If $X_1,X_2,\ldots,X_n$ are jointly normal random variables, then every linear combination

$$
a + \sum_{i=1}^{n} a_iX_i
$$

is also normally distributed.
:::

This includes linear transformations of a single normal random variable. If

$$
X \sim \mathcal{N}(\mu,\sigma^2)
$$

then standardizing $X$ gives

$$
Z := \frac{X-\mu}{\sigma} \sim \mathcal{N}(0,1)
$$

Conversely, if $Z \sim \mathcal{N}(0,1)$, then

$$
X := \sigma Z+\mu \sim \mathcal{N}(\mu,\sigma^2)
$$

### The Standard Normal Distribution {#sec-standard-normal}

For a normal random variable 
$X \sim \mathcal{N}(\mu,\sigma^2)$
the probability density function is

$$
f_X(x)
=
\frac{1}{\sqrt{2\pi\sigma^2}}
\exp\!\left\{-\frac{(x-\mu)^2}{2\sigma^2}\right\}
$$

::: {#def-standard-normal}
## Standard Normal Distribution

A standard normal random variable has distribution

$$
Z \sim \mathcal{N}(0,1)
$$

and probability density function

$$
f_Z(z)
=
\frac{1}{\sqrt{2\pi}}
\exp\!\left\{-\frac{z^2}{2}\right\}
$$
:::

The normal cumulative distribution function cannot be written in elementary closed form, but it can be evaluated numerically.

::: {.callout-note}
## Notation

For $Z \sim \mathcal{N}(0,1)$, the standard normal cumulative distribution function is

$$
\Phi(z) = P(Z \leq z)
$$
:::

In R, standard normal probabilities are computed using `pnorm()`.

```{r}
pnorm(1.23)
```

This gives approximately $0.891$.

::: {#exm-normal-probability}
## Normal Probability

Suppose that

$$
X \sim \mathcal{N}(1,4)
$$

Find $P(X \leq 2)$.

Because $\mu=1$ and $\sigma=\sqrt{4}=2$,

$$
\begin{aligned}
P(X \leq 2)
&=
P\!\left(
\frac{X-\mu}{\sigma}
\leq
\frac{2-1}{\sqrt{4}}
\right)
&& \text{standardize }X \\
&=
P(Z \leq 0.5) \\
&=
\Phi(0.5) \\
&\approx 0.6915
\end{aligned}
$$

```{r}
pnorm(0.5)
```
:::

::: {#exm-normal-sample-mean}
## Sample Mean of Normal Random Variables

Suppose that $X_1,X_2,X_3$ are independent and

$$
X_i \sim \mathcal{N}(1,4)
$$

Find $P(\overline{X} \leq 2)$.

Because a linear combination of independent normal random variables is normal, $\overline{X}$ is normally distributed. Its mean and variance are

$$
\begin{aligned}
\mathbb{E}[\overline{X}]
&= \mathbb{E}[X_1] \\
&= 1 \\
\operatorname{Var}(\overline{X})
&= \frac{\operatorname{Var}(X_1)}{n} \\
&= \frac{4}{3}
\end{aligned}
$$

Therefore,

$$
\overline{X} \sim \mathcal{N}\!\left(1,\frac{4}{3}\right)
$$

and

$$
\begin{aligned}
P(\overline{X} \leq 2)
&=
P\!\left(
\frac{\overline{X}-\mu_{\overline{X}}}
{\sigma_{\overline{X}}}
\leq
\frac{2-1}{2/\sqrt{3}}
\right)
&& \text{standardize }\overline{X} \\
&=
P\!\left(Z \leq \frac{\sqrt{3}}{2}\right) \\
&=
\Phi\!\left(\frac{\sqrt{3}}{2}\right) \\
&\approx 0.8068
\end{aligned}
$$

```{r}
pnorm(sqrt(3) / 2)
```
:::

## Convergence in Distribution {#sec-convergence-distribution}

::: {#def-convergence-distribution}
## Convergence in Distribution

Let $X_1,X_2,X_3,\ldots$ be a sequence of random variables, where $X_n$ has cumulative distribution function

$$
F_n(x) = P(X_n \leq x)
$$

Let $X$ be a random variable with cumulative distribution function

$$
F(x) = P(X \leq x)
$$

The sequence $X_1,X_2,X_3,\ldots$ converges in distribution to $X$ if

$$
\lim_{n\to\infty}F_n(x) = F(x)
$$

at every point $x$ at which $F$ is continuous.

We write

$$
X_n \xrightarrow{d} X
$$
:::

Convergence in distribution is a weak form of convergence. The random variables themselves need not become close to one another; rather, their distributions become close.

For example, if

$$
X_n \sim \mathcal{N}\!\left(\frac{1}{n},\sigma^2\right)
$$

then

$$
X_n \xrightarrow{d} X
\qquad
\text{where}
\qquad
X \sim \mathcal{N}(0,\sigma^2)
$$

In particular, when $\sigma^2=1$,

$$
X_n \xrightarrow{d} Z
\qquad
\text{where}
\qquad
Z \sim \mathcal{N}(0,1)
$$

## Sampling Distributions of Exponential Sample Means {#sec-exponential-sample-means}

Suppose that

$$
X \sim \operatorname{Exp}(\operatorname{rate}=\lambda)
$$

The slides compare $10{,}000$ sample means for sample sizes

$$
n \in \{1,2,3,4,5,10,100,1000,10000\}
$$

As $n$ increases, the sampling distribution of $\overline{X}_n$ becomes increasingly symmetric and concentrated around

$$
\mathbb{E}[X] = \frac{1}{\lambda}
$$

For exponential random variables,

$$
\sum_{i=1}^{n}X_i \sim \operatorname{Gamma}(n,\operatorname{rate}=\lambda)
$$

so the sample mean has the exact distribution

$$
\overline{X}_n
\sim
\operatorname{Gamma}(n,\operatorname{rate}=n\lambda)
$$

The following R code reproduces the sequence of histograms efficiently.

```{r}
#| fig-width: 12
#| fig-height: 10
#| fig-cap: "Sampling distributions of 10,000 exponential sample means"

set.seed(123)

B <- 10000
lambda <- 1
sample_sizes <- c(1, 2, 3, 4, 5, 10, 100, 1000, 10000)

old_par <- par(mfrow = c(3, 3), mar = c(3, 3, 2, 1))

for (n in sample_sizes) {
  x_bar <- rgamma(B, shape = n, rate = n * lambda)

  hist(
    x_bar,
    probability = TRUE,
    breaks = 16,
    main = paste("n =", n),
    xlab = expression(bar(X)),
    ylab = "Density"
  )
}

par(old_par)
```

## The Central Limit Theorem {#sec-central-limit-theorem}

::: {#thm-central-limit-theorem}
## Central Limit Theorem

Let $X_1,X_2,X_3,\ldots$ be independent and identically distributed random variables from any distribution with finite mean and variance

$$
\mathbb{E}[X_i] = \mu
\qquad
\operatorname{Var}(X_i) = \sigma^2 < \infty
$$

Define the sample mean

$$
\overline{X}_n
=
\frac{1}{n}\sum_{i=1}^{n}X_i
$$

Then

$$
\frac{\overline{X}_n-\mu}{\sigma/\sqrt{n}}
\xrightarrow{d}
\mathcal{N}(0,1)
$$
:::

## Asymptotic Normality {#sec-asymptotic-normality}

::: {#def-asymptotic-normality}
## Asymptotically Normal Random Variables

A sequence of random variables $X_n$ is asymptotically normal if there exist sequences of real numbers $\{a_n\}$ and positive real numbers $\{b_n\}$ such that

$$
\frac{X_n-a_n}{\sqrt{b_n}}
\xrightarrow{d}
\mathcal{N}(0,1)
$$

We write

$$
X_n
\overset{\mathrm{asymp}}{\sim}
\mathcal{N}(a_n,b_n)
$$
:::

This notation does not mean that $X_n$ has an exact normal distribution. It means that its standardized distribution converges to the standard normal distribution.

The Central Limit Theorem therefore states that the sample mean is asymptotically normal:

$$
\overline{X}_n
\overset{\mathrm{asymp}}{\sim}
\mathcal{N}\!\left(\mu,\frac{\sigma^2}{n}\right)
$$

::: {#exm-clt-gamma-sample-mean}
## Gamma Sample Mean

Let $\overline{X}$ be the sample mean of a random sample of size $n=100$ from a gamma distribution with shape $3$ and rate $2$:

$$
X_i \sim \operatorname{Gamma}(3,\operatorname{rate}=2)
$$

The population mean and variance are

$$
\begin{aligned}
\mathbb{E}[X_i]
&=
\frac{3}{2} \\
\operatorname{Var}(X_i)
&=
\frac{3}{2^2} \\
&=
\frac{3}{4}
\end{aligned}
$$

Therefore, the sample mean has

$$
\begin{aligned}
\mathbb{E}[\overline{X}]
&=
\mathbb{E}[X_i] \\
&=
\frac{3}{2} \\
\operatorname{Var}(\overline{X})
&=
\frac{\operatorname{Var}(X_i)}{n} \\
&=
\frac{3/4}{100} \\
&=
\frac{3}{400}
\end{aligned}
$$

By the Central Limit Theorem,

$$
\overline{X}
\overset{\mathrm{asymp}}{\sim}
\mathcal{N}\!\left(\frac{3}{2},\frac{3}{400}\right)
$$

::: {.callout-warning}
## Threshold Used in the Slides

The question states $P(\overline{X}>1.49)$, but the subsequent derivation uses $1.4$. The reported probability $0.8749$ corresponds to the threshold $1.4$.
:::

For the threshold $1.4$,

$$
\begin{aligned}
P(\overline{X}>1.4)
&\approx
P\!\left(
\frac{\overline{X}-\mu_{\overline{X}}}
{\sigma_{\overline{X}}}
>
\frac{1.4-1.5}{\sqrt{3/400}}
\right)
&& \text{apply the normal approximation} \\
&=
P(Z>-1.1547) \\
&=
1-P(Z\leq -1.1547) \\
&=
1-\Phi(-1.1547) \\
&\approx
0.8749
\end{aligned}
$$

The calculation shown in the slides is obtained in R using

```{r}
1 - pnorm(-1.15)
```

For the threshold $1.49$ stated in the question,

$$
\begin{aligned}
P(\overline{X}>1.49)
&\approx
P\!\left(
Z>
\frac{1.49-1.5}{\sqrt{3/400}}
\right) \\
&=
P(Z>-0.1155) \\
&\approx
0.5460
\end{aligned}
$$

```{r}
1 - pnorm((1.49 - 1.5) / sqrt(3 / 400))
```
:::

::: {.callout-note}
## Standardizing the Sample Mean

Because

$$
\mu_{\overline{X}}=\mu
\qquad
\sigma_{\overline{X}}=\frac{\sigma}{\sqrt{n}}
$$

we have

$$
\frac{
\overline{X}-\underbrace{\mu_{\overline{X}}}_{\mu}
}{
\underbrace{\sigma_{\overline{X}}}_{\sigma/\sqrt{n}}
}
=
\frac{\overline{X}-\mu}{\sigma/\sqrt{n}}
$$
:::

## Reflection

