---
title: "Lesson 2: The t- and Chi-Squared Distributions and the Sample Variance"
subtitle: "Statistical Estimation for Data Science and AI"
---

::: {.column-margin}
{{< video 01_the-chi-squared-and-t-distributions.mp4 >}}

[Transcript](01_the-chi-squared-and-t-distributions.en.txt)
[Handout](02_video-slides-for-the-chi-squared-and-t-distributions_4.2_Lesson_Slides.pdf){width=700 height=980}
:::

## Slides on the Chi-Squared Distribution {#sec-chi-squared-distribution}

::: {#def-chi-squared-distribution}
## Chi-Squared Distribution

A chi-squared RV with $n$ degrees of freedom is the special case of a gamma RV.
$$
X \sim \Gamma\!\left(\frac{n}{2},\frac{1}{2}\right)
$$

where the gamma distribution is parameterized by shape and rate

Its probability density function is

$$
f_X(x) = \frac{1}{\Gamma(n/2)} \left(\frac{1}{2}\right)^{n/2}
x^{n/2-1} e^{-x/2} \mathbb{I}_{(0,\infty)}(x)
$$

We write

$$
X \sim \chi^2(n)
$$
:::

The parameter $n$ is called the number of degrees of freedom

The mean and variance are

$$
\begin{aligned}
\mathbb{E}[X] &= n \\
\operatorname{Var}(X) &= 2n
\end{aligned}
$$

### Shape of the Density {#sec-chi-squared-density}

The chi-squared density is supported on the positive real line. 
For small $n$ it is strongly right-skewed

```{r}
#| label: fig-chi-squared-density
#| fig-cap: "Density of a chi-squared distribution with one degree of freedom"
#| echo: false
#| warning: false

x <- seq(0.001, 1, length.out = 1000)
plot(
  x,
  dchisq(x, df = 1),
  type = "l",
  xlab = "x",
  ylab = expression(f[X](x)),
  ylim = c(0, 12)
)
```

## Things About the Chi-Squared Distribution {#sec-chi-squared-properties}

### Moment-Generating Function {#sec-chi-squared-mgf}

For a gamma random variable parameterized by shape $\alpha$ and rate $\beta$,

$$
X \sim \Gamma(\alpha,\beta)
$$

its moment-generating function is

$$
M_X(t)
=
\left(\frac{\beta}{\beta-t}\right)^\alpha
\qquad t<\beta
$$

Since

$$
X \sim \chi^2(n)
=
\Gamma\!\left(\frac{n}{2},\frac{1}{2}\right)
$$

we obtain

$$
\begin{aligned}
M_X(t)
&=
\left(
\frac{1/2}{1/2-t}
\right)^{n/2} \\
&=
(1-2t)^{-n/2}
\qquad t<\frac12
\end{aligned}
$$

::: {#thm-sum-independent-chi-squared}
## Sum of Independent Chi-Squared Random Variables

Suppose that $X_1,X_2,\ldots,X_k$ are independent and

$$
X_i \sim \chi^2(n_i)
$$

Then

$$
Y=\sum_{i=1}^k X_i
$$

has distribution

$$
Y \sim \chi^2\!\left(\sum_{i=1}^k n_i\right)
$$
:::

The result follows from the moment-generating function

$$
\begin{aligned}
M_Y(t)
&=
\mathbb{E}\!\left[e^{tY}\right] \\
&=
\mathbb{E}\!\left[
 e^{t\sum_{i=1}^k X_i}
\right] \\
&=
\mathbb{E}\!\left[
 \prod_{i=1}^k e^{tX_i}
\right] \\
&=
\prod_{i=1}^k
\mathbb{E}\!\left[e^{tX_i}\right]
&& \text{by independence} \\
&=
\prod_{i=1}^k M_{X_i}(t) \\
&=
\prod_{i=1}^k
\left(
\frac{1/2}{1/2-t}
\right)^{n_i/2} \\
&=
\left(
\frac{1/2}{1/2-t}
\right)^{\frac12\sum_{i=1}^k n_i}
\end{aligned}
$$

Thus the sum of independent chi-squared random variables is itself chi-squared, and their degrees of freedom add

## The Chi-Squared--Normal Relationship {#sec-chi-squared-normal-relationship}

::: {#thm-squared-standard-normal}
## Square of a Standard Normal Random Variable

Let

$$
X \sim \mathcal{N}(0,1)
$$

and define

$$
Y=X^2
$$

Then

$$
Y \sim \chi^2(1)
$$
:::

This result can be shown either with moment-generating functions or by transforming the normal density

For $y>0$, the transformation $y=x^2$ has the two inverse branches $x=\sqrt y$ and $x=-\sqrt y$. Therefore

$$
\begin{aligned}
f_Y(y)
&=
f_X(\sqrt y)
\left|\frac{d}{dy}\sqrt y\right|
+
f_X(-\sqrt y)
\left|\frac{d}{dy}(-\sqrt y)\right| \\
&=
\frac{f_X(\sqrt y)+f_X(-\sqrt y)}{2\sqrt y} \\
&=
\frac{1}{\sqrt{2\pi}}
y^{-1/2}e^{-y/2}
\mathbb{I}_{(0,\infty)}(y) \\
&=
\frac{1}{\Gamma(1/2)}
\left(\frac12\right)^{1/2}
y^{1/2-1}e^{-y/2}
\mathbb{I}_{(0,\infty)}(y)
\end{aligned}
$$

This is the density of $\chi^2(1)$

### Transformation Formula {#sec-transformation-formula}

For a one-to-one transformation $Y=g(X)$ with inverse $X=g^{-1}(Y)$,

$$
f_Y(y)
=
f_X\!\left(g^{-1}(y)\right)
\left|
\frac{d}{dy}g^{-1}(y)
\right|
$$

For a bivariate transformation

$$
\begin{aligned}
Y_1 &= g_1(X_1,X_2) \\
Y_2 &= g_2(X_1,X_2)
\end{aligned}
$$

write the inverse transformation as

$$
\begin{aligned}
X_1 &= h_1(Y_1,Y_2) \\
X_2 &= h_2(Y_1,Y_2)
\end{aligned}
$$

Then

$$
f_{Y_1,Y_2}(y_1,y_2)
=
f_{X_1,X_2}\!\left(
 h_1(y_1,y_2),
 h_2(y_1,y_2)
\right)
\left|J\right|
$$

where

$$
J
=
\begin{vmatrix}
\dfrac{\partial x_1}{\partial y_1}
&
\dfrac{\partial x_1}{\partial y_2}
\\[6pt]
\dfrac{\partial x_2}{\partial y_1}
&
\dfrac{\partial x_2}{\partial y_2}
\end{vmatrix}
$$

## The t-Distribution {#sec-t-distribution}

::: {#def-t-distribution}
## t-Distribution

Let

$$
Z \sim \mathcal{N}(0,1)
$$

and

$$
W \sim \chi^2(n)
$$

be independent random variables. Define

$$
T
=
\frac{Z}{\sqrt{W/n}}
$$

Then $T$ has a t-distribution with $n$ degrees of freedom, written

$$
T \sim t(n)
$$
:::

### Density from a Jacobian Transformation {#sec-t-density-jacobian}

Take

$$
\begin{aligned}
Y_1
&=
\frac{X_1}{\sqrt{X_2/n}}
=
T \\
Y_2
&=
X_2
\end{aligned}
$$

where $X_1=Z$ and $X_2=W$. The inverse transformation is

$$
\begin{aligned}
X_1
&=
y_1\sqrt{\frac{y_2}{n}} \\
X_2
&=
y_2
\end{aligned}
$$

with Jacobian

$$
\left|J\right|
=
\sqrt{\frac{y_2}{n}}
$$

Therefore

$$
\begin{aligned}
f_T(t)
&=
\int_0^\infty
f_Z\!\left(t\sqrt{\frac{w}{n}}\right)
f_W(w)
\sqrt{\frac{w}{n}}
\,dw \\
&=
\frac{1}{\sqrt{2\pi}}
\frac{1}{2^{n/2}\Gamma(n/2)}
\frac{1}{\sqrt n}
\int_0^\infty
w^{(n+1)/2-1}
\exp\!\left\{
-\frac{w}{2}
\left(1+\frac{t^2}{n}\right)
\right\}
\,dw \\
&=
\frac{\Gamma\!\left(\frac{n+1}{2}\right)}
     {\Gamma\!\left(\frac n2\right)\sqrt{n\pi}}
\left(1+\frac{t^2}{n}\right)^{-(n+1)/2}
\end{aligned}
$$

Thus the t-density is

$$
f_T(t)
=
\frac{\Gamma\!\left(\frac{n+1}{2}\right)}
     {\Gamma\!\left(\frac n2\right)\sqrt{n\pi}}
\left(1+\frac{t^2}{n}\right)^{-(n+1)/2}
\qquad -\infty<t<\infty
$$

Its mean and variance are

$$
\begin{aligned}
\mathbb{E}[T] &= 0 && n>1 \\
\operatorname{Var}(T) &= \frac{n}{n-2} && n>2
\end{aligned}
$$

::: {.callout-note}
## Moment Existence

For $n\leq 1$, the mean does not exist. For $1<n\leq 2$, the mean exists but the variance is infinite
:::

### Comparison with the Standard Normal Distribution {#sec-t-normal-comparison}

The t-distribution is symmetric about zero but has heavier tails than the standard normal distribution. As the degrees of freedom increase, it approaches $\mathcal{N}(0,1)$

```{r}
#| label: fig-t-normal-comparison
#| fig-cap: "The t-distribution approaches the standard normal distribution as its degrees of freedom increase"
#| echo: false
#| warning: false

x <- seq(-4, 4, length.out = 1000)
plot(
  x,
  dnorm(x),
  type = "l",
  lwd = 2,
  col = "navy",
  xlab = "t",
  ylab = "density",
  ylim = c(0, 0.42)
)
lines(x, dt(x, df = 20), lwd = 2, col = "firebrick")
lines(x, dt(x, df = 10), lwd = 2, col = "darkgreen")
legend(
  "topright",
  legend = c("N(0,1)", "t(20)", "t(10)"),
  col = c("navy", "firebrick", "darkgreen"),
  lwd = 2,
  bty = "n"
)
```

## Why These Distributions Matter {#sec-why-chi-squared-and-t}

Suppose

$$
X_1,X_2,\ldots,X_n
\stackrel{\mathrm{iid}}{\sim}
\mathcal{N}(\mu,\sigma^2)
$$

Define the sample mean and sample variance by

$$
\bar X
=
\frac1n\sum_{i=1}^n X_i
$$

and

$$
S^2
=
\frac{1}{n-1}
\sum_{i=1}^n
(X_i-\bar X)^2
$$

### Decomposition of the Total Squared Deviation {#sec-squared-deviation-decomposition}

Write

$$
X_i-\mu
=
(X_i-\bar X)+(\bar X-\mu)
$$

Then

$$
\begin{aligned}
\sum_{i=1}^n (X_i-\mu)^2
&=
\sum_{i=1}^n
\left[(X_i-\bar X)+(\bar X-\mu)\right]^2 \\
&=
\sum_{i=1}^n (X_i-\bar X)^2
+2(\bar X-\mu)
\sum_{i=1}^n (X_i-\bar X)
+n(\bar X-\mu)^2 \\
&=
\sum_{i=1}^n (X_i-\bar X)^2
+n(\bar X-\mu)^2
&& \sum_{i=1}^n(X_i-\bar X)=0
\end{aligned}
$$

Indeed,

$$
\begin{aligned}
\sum_{i=1}^n(X_i-\bar X)
&=
\sum_{i=1}^nX_i-n\bar X \\
&=
\sum_{i=1}^nX_i
-n\left(\frac1n\sum_{i=1}^nX_i\right) \\
&=0
\end{aligned}
$$

Dividing the decomposition by $\sigma^2$ gives

$$
\underbrace{
\sum_{i=1}^n
\left(\frac{X_i-\mu}{\sigma}\right)^2
}_{\color{blue}{Y_1}}
=
\underbrace{
\frac{\sum_{i=1}^n(X_i-\bar X)^2}{\sigma^2}
}_{\color{red}{Y_2}}
+
\underbrace{
\left(
\frac{\bar X-\mu}{\sigma/\sqrt n}
\right)^2
}_{\color{green}{Y_3}}
$$

or equivalently

$$
Y_1=Y_2+Y_3
$$

where

$$
\begin{aligned}
Y_1
&=
\sum_{i=1}^n
\left(\frac{X_i-\mu}{\sigma}\right)^2 \\
Y_2
&=
\frac{(n-1)S^2}{\sigma^2} \\
Y_3
&=
\left(
\frac{\bar X-\mu}{\sigma/\sqrt n}
\right)^2
\end{aligned}
$$

### Distribution of the Components {#sec-normal-sample-components}

Since

$$
\frac{X_i-\mu}{\sigma}
\sim
\mathcal{N}(0,1)
$$

we have

$$
\left(
\frac{X_i-\mu}{\sigma}
\right)^2
\sim
\chi^2(1)
$$

and therefore

$$
Y_1
=
\sum_{i=1}^n
\left(
\frac{X_i-\mu}{\sigma}
\right)^2
\sim
\chi^2(n)
$$

Also,

$$
\frac{\bar X-\mu}{\sigma/\sqrt n}
\sim
\mathcal{N}(0,1)
$$

so

$$
Y_3
=
\left(
\frac{\bar X-\mu}{\sigma/\sqrt n}
\right)^2
\sim
\chi^2(1)
$$

For a normal sample, $\bar X$ and $S^2$ are independent. Consequently, $Y_2$ and $Y_3$ are independent

### Distribution of the Sample Variance {#sec-sample-variance-chi-squared}

Because $Y_1=Y_2+Y_3$ and $Y_2$ and $Y_3$ are independent,

$$
\begin{aligned}
M_{Y_1}(t)
&=
M_{Y_2+Y_3}(t) \\
&=
M_{Y_2}(t)M_{Y_3}(t)
\end{aligned}
$$

Hence

$$
\begin{aligned}
M_{Y_2}(t)
&=
\frac{M_{Y_1}(t)}{M_{Y_3}(t)} \\
&=
\frac{
\left(\dfrac{1/2}{1/2-t}\right)^{n/2}
}{
\left(\dfrac{1/2}{1/2-t}\right)^{1/2}
} \\
&=
\left(
\frac{1/2}{1/2-t}
\right)^{(n-1)/2}
\end{aligned}
$$

This is the moment-generating function of a chi-squared random variable with $n-1$ degrees of freedom

::: {#thm-sample-variance-chi-squared}
## Chi-Squared Distribution of the Sample Variance

For

$$
X_1,X_2,\ldots,X_n
\stackrel{\mathrm{iid}}{\sim}
\mathcal{N}(\mu,\sigma^2)
$$

we have

$$
\frac{(n-1)S^2}{\sigma^2}
\sim
\chi^2(n-1)
$$
:::

### The Studentized Sample Mean {#sec-studentized-sample-mean}

Let

$$
Z
=
\frac{\bar X-\mu}{\sigma/\sqrt n}
\sim
\mathcal{N}(0,1)
$$

and

$$
W
=
\frac{(n-1)S^2}{\sigma^2}
\sim
\chi^2(n-1)
$$

Since $Z$ and $W$ are independent,

$$
\begin{aligned}
\frac{Z}{\sqrt{W/(n-1)}}
&=
\frac{
\dfrac{\bar X-\mu}{\sigma/\sqrt n}
}{
\sqrt{
\dfrac{(n-1)S^2/\sigma^2}{n-1}
}
} \\
&=
\frac{
\dfrac{\bar X-\mu}{\sigma/\sqrt n}
}{S/\sigma} \\
&=
\frac{\bar X-\mu}{S/\sqrt n}
\end{aligned}
$$

Therefore

$$
\frac{\bar X-\mu}{S/\sqrt n}
\sim
t(n-1)
$$

## Reflection

