
#### Probability, Expectation, and Variance"



::: {#exr-01-refrigerator }
## Expected refrigerator price
[refrigerator]{.column-margin}

A big box store sells three different models of refrigerators having 
20, 21.5, and 26.8 cubic feet of usable storage space, 
respectively. Let X = the amount of storage space purchased by the next 
customer to buy a freezer. Suppose that X has probability mass function given by

| x (cubic feet) | 20  | 21.5 | 26.8 |
|----------------|-----|------|------|
| $\mathbb{P}(x)$| 0.1 | 0.7  | 0.2  |
: PMF for refrigerator storage space {#tbl-refrigerator-pmf }

If the price of a refrigerator with $X$ cubic feet of capacity is 
$PRICE=(20X+100)$, what is the expected price paid by the next customer to buy a freezer?
:::

::: {.callout-tip}
## Solution

1. Solution Strategy: 
    1. we need to find the expected value of PRICE
    2. PRICE is a linear transformation of the X variable
    3. The expectation can then be calculated as usual ie 
       $$
       \sum_i P(x_i) \cdot PRICE(x_i)
       $$ {#eq-refrigerator-expected-price }

```{r}

data <- data.frame(
  x   = c(20,21.5,26.8),
  p_x = c(0.1, 0.7, 0.2)
)
data


Expecred_Price <- function(data) {
  sum((20*data$x+100) * data$p_x) #implementation of @eq-refrigerator-expected-price
}
Expecred_Price(data)
```

:::


::: {#exr-02-circuit-board }
## Circuit board lifetimes
[Circuit board]{.column-margin} 

In proof testing of circuit boards, the probability that any particular diode will fail is 0.03 . Suppose a circuit board contains 150 diodes and that each one performs or fails independently of all others.

- [How many diodes would you expect to fail]{.mark}, and 
- [what is the standard deviation]{.mark} of the number that are expected to fail?

:::


::: {.callout-tip}
## Solution

1. Solution Strategy: 
    1. Let $X$ = number of diodes that fail. Since each diode is an iid Bernoulli trials with fixed probability of failure and independence. Their sum will follow a Binomial distribution.
    2. $\therefore X \sim \mathcal{Bin}(n = 150, p = 0.03)$
    3. The expectation and standard deviation of a binomial distribution are given by:
       - $E(X) = n \cdot p \qquad SD(X) = \sqrt{n \cdot p \cdot (1 - p)}$

```{r}
n <- 150
p <- 0.03
expected_failures <- n * p
sd_failures <- sqrt(n * p * (1 - p))
expected_failures
sd_failures
```
:::


---



::: {#exr-03-skink-dispersal }
## Skink dispersal distances
[Skink dispersal]{.column-margin}

Let $X$ denote the distance, in meters, that an animal moves from its birth site to the first territorial vacancy it encounters. 
Suppose that for Solomon Island prehensile-tailed skinks, $X$ has an exponential distribution with rate parameter  $\lambda = 0.014$.

What is the probability that the distance exceeds the mean distance by more than 2 standard deviations?
:::


::: {.callout-caution collapse="true"}
## Check Parameterization
The best way to mess up is to get the parameterization of the exponential distribution wrong. 
Even worse if we also have a Poisson in the mix. 
Jem cautions us about this several times.
i.e. do we need to use $\frac{1}{\lambda}$?)

PDF from handout:
$$
f(x) = \lambda e^{-\lambda x} \mathbb{I}_{x \geq 0}(x)
$$

Does R parameterize the same way?
:::

::: {.callout-tip collapse="true"}
## Solution

- strategy:
    1. Given: $X \sim \mathcal{Exp}(rate = 0.014)$  
    2. We have $\mu=\frac{1}{\lambda}$ and $\text{sd}=\frac{1}{\lambda}$  from wikipedia

```{r}
rate<-0.014
pexp(1/rate + 2*(1/rate), rate = rate, lower.tail = FALSE)
```
:::


::: {#exr-04-mode-gamma }
## Mode of gamma distribution
[Mode of gamma distribution]{.column-margin}

[The mode of a continuous distribution with probability density function $f(x)$ is the value of $x$ that maximizes $f(x)$.]{.mark}

What is the mode of the  gamma distribution with parameters $\alpha$ and 
$\beta$ for $\alpha > 1$?
:::

::: {.callout-tip}
## Solution 
- Strategy:
    1. The PDF of a gamma distribution is given by:
       $$
       f(x) = \frac{1}{\Gamma(\alpha) }\beta^\alpha x^{\alpha - 1} e^{-x\beta} \mathbb{I}_{x \geq 0}(x)
       $$
    2. To find the mode, we take the derivative of $f(x)$ with respect to $x$, set it to zero, and solve for $x$.
    3. The derivative is:

       $$
       f'(x) = \frac{1}{\Gamma(\alpha) }\beta^\alpha \left( (\alpha - 1)x^{\alpha - 2} e^{-x\beta} - \beta x^{\alpha - 1} e^{-x\beta} \right)
       $$

    4. Setting $f'(x) = 0$ and solving for $x$ gives:
       $$
       (\alpha - 1) \cancel{x^{\alpha - 2}} \cancel{e^{-x \beta}} = \beta x^{\cancel{\alpha - 1}} \cancel{e^{-x\beta}}
       $$

       Simplifying, we find:
       
       $$
       x = \frac{(\alpha - 1)}{\beta}
       $$

    5. Therefore, the mode of the gamma distribution for $\alpha > 1$ is:
       $$
       \text{Mode} = \frac{(\alpha - 1)}{\beta}
       $$
:::


::: {#exr-05-poisson-second-moment }
## Poisson second moment
[Poisson second moment]{.column-margin}

Suppose that $X$ is a Poisson random variable with parameter 
$\lambda$. What is  $\mathbb{E}[X^2]$?
:::

::: {.callout-tip}
## Solution
- Strategy:
    1. We know that for a Poisson distribution, $E[X] = \lambda$ and $Var(X) = \lambda$.
    2. We can use the relationship between variance and expectation:
       $$
       \text{Var}(X) = \mathbb{E}[X^2] - (\mathbb{E}[X])^2
       $$
    3. Rearranging gives:
       $$
       \mathbb{E}[X^2] = \text{Var}(X) + (\mathbb{E}[X])^2
       $$
    4. Substituting the known values:
       $$
       \mathbb{E}[X^2] = \lambda + \lambda^2
       $$
:::