---
title: "Lab: Point Estimation"
subtitle: "Statistical Estimation for Data Science and AI"
---

```{r}
#| deletable: false
#| editable: false
#| nbgrader: {cell_type: code, checksum: a7e543e438ae35de9e51cb1d2556db56, grade: false, grade_id: cell-81d3dcb436100e63, locked: true, schema_version: 3, solution: false, task: false}
#| vscode: {languageId: r}
# Load Necesary Libraries for Autograding
library(testthat)
```

## Problem 1

Let the joint PDF $f(X,Y)$ be represented by the following table of values:


|  | y=-1| y=0 | y=1 | y=2 |
| :--- | :----: | :---: | :---: | ---: |
| $x=2$ | 0.1 | 0 | 0.15 | 0 |
| $x=3$ | 0.1 | 0.1 | 0.05 | 0.05 |
| $x=4$ | 0 | 0.05 | 0.15 | 0.25 |

 **Part A)** Find $E[X]$ and $E[Y]$. Save your answers as `p1.a.ex` and `p1.a.ey`.

```{r}
#| deletable: false
#| nbgrader: {cell_type: code, checksum: ebff4591302576a2f85832c578415fc8, grade: false, grade_id: cell-c01546f6e409fcc4, locked: false, schema_version: 3, solution: true, task: false}
#| vscode: {languageId: r}
p1.a.ex = NA
p1.a.ey = NA

# your code here
# to estimate E[X] we need to margenelize y.
# to estimate E[Y] we need to margenelize x.
# the data should be in a matrix or a df or a tibble
library(tidyverse)
#library(tibble) 
#library(dplyr) 
#library(tidyr)
my_data <- tribble(
   ~x,  ~y,  ~p,
    2,  -1,  0.1,
    3,  -1,  0.1,
    4,  -1,  0,    
    2,   0,  0,
    3,   0,  0.1,
    4,   0,  0.05,    
    2,   1,  0.15,
    3,   1,  0.05,
    4,   1,  0.15,    
    2,   2,  0,
    3,   2,  0.05,
    4,   2,  0.25
)
my_data

## lets recreate the table (to check)
my_data_wide <- my_data %>%
  pivot_wider(names_from = y, values_from = p, names_prefix = "y=")

my_data_wide

## this doesn't have the y values
## this shows the main difference between df or tibbles and spreadsheet style data
## in spreadsheets the row and column headers are part of the data
## in df or tibbles the row and column headers are meta data

## tip: when we need to store intermediate result we will use summerise to create a new tibble

## We need to margielise y out than agregate x* marginal_p
my_data %>% 
    group_by(x) %>% # to mararginalize out y we need to sum y by X=x
    summarize(marginal_px = sum(p)) %>% # df for marginal X dist
    mutate(E_X_component = x * marginal_px) %>% # col for x * marginal_px
    summarize(E_X = sum(E_X_component)) %>% # df for agg E[X]
    pull(E_X) -> p1.a.ex # extact the result

p1.a.ex

my_data %>%
  group_by(y) %>% # to mararginalize out x we need to sum y by X=x
  summarize(marginal_py = sum(p)) %>% # df for marginal Y dist
  mutate(E_Y_component = y * marginal_py) %>% # col for y * marginal_py
  summarize(E_Y = sum(E_Y_component)) %>% # df for agg E[Y]
  pull(E_Y) -> p1.a.ey # extact the result

p1.a.ey
```

```{r}
#| deletable: false
#| editable: false
#| nbgrader: {cell_type: code, checksum: 78fb7ab65d8dc28608552b1227d972fe, grade: true, grade_id: cell-5b1889eb8efdf2a8, locked: true, points: 1, schema_version: 3, solution: false, task: false}
#| vscode: {languageId: r}
# Hidden Test Cell
# NOTE: This cell contains hidden tests. You will not see if you passed these tests until you submit your assignment.
# Any cell labeled "Hidden Test Cell" MAY have hidden tests.
```

**Part B)** Find $E[XY]$. Save your answer as `p1.b.exy`.

```{r}
#| deletable: false
#| nbgrader: {cell_type: code, checksum: 41d0ffd93d95eef62309e2114fb86b98, grade: false, grade_id: cell-5c78993b7207eb33, locked: false, schema_version: 3, solution: true, task: false}
#| vscode: {languageId: r}
p1.b.exy = NA

# your code here
my_data %>%
  mutate(XY_component = x * y * p)    %>% # add col for x*y*p
  summarize(E_XY = sum(XY_component)) %>% # sore its sum in a new df
  pull(E_XY) -> p1.b.exy                  # assign to variable 
```

```{r}
#| deletable: false
#| editable: false
#| nbgrader: {cell_type: code, checksum: a23d003bd351d468bac2f4fad6a94263, grade: true, grade_id: cell-a07fd01df677d2f4, locked: true, points: 1, schema_version: 3, solution: false, task: false}
#| vscode: {languageId: r}
# Hidden Test Cell
```

**Part C)** Are $X$ and $Y$ independent? Answer the boolean value `TRUE` if they are independent and `FALSE` if they are not independent. Save your answer as `p1.c`.

```{r}
#| deletable: false
#| nbgrader: {cell_type: code, checksum: fe14614bf3cce08e14aed7f4b4a25248, grade: false, grade_id: cell-7adb1551e02e131a, locked: false, schema_version: 3, solution: true, task: false}
#| vscode: {languageId: r}
p1.c = NA

# your code here
# if x and y are independent then e[xy]=e[x]*e[y]
p1.c = p1.b.exy == p1.a.ex*p1.a.ey 
p1.c
```

```{r}
#| deletable: false
#| editable: false
#| nbgrader: {cell_type: code, checksum: 204399f6410f78b4c72282f8c4075223, grade: true, grade_id: cell-366a6f3eed5323a8, locked: true, points: 1, schema_version: 3, solution: false, task: false}
#| vscode: {languageId: r}
# Hidden Test Cell
if(!test_that("Check that answer is correct type", expect_is(p1.c, "logical"))){
    print("Make sure your answer is a boolean!")
}
```

**Part D)** Define $U=X+Y$. What is $P(U \ge 4)$? Save your answer as `p1.d`.

```{r}
#| deletable: false
#| nbgrader: {cell_type: code, checksum: fff08d5b3bcb2bfcd459b229f6cc733f, grade: false, grade_id: cell-3fe9def7644611b9, locked: false, schema_version: 3, solution: true, task: false}
#| vscode: {languageId: r}
p1.d = NA

# your code here
my_data %>%
  mutate(u = x + y) %>%               # create rv u
  filter(u >= 4) %>%                  # filter the event
  summarise(probability = sum(p)) %>% # sum of p into new df 
  pull(probability) -> p1.d           # assign into result

p1.d
```

```{r}
#| deletable: false
#| editable: false
#| nbgrader: {cell_type: code, checksum: 15b37e977aefe7911d39b5be5923234d, grade: true, grade_id: cell-51a8c559d0bcea8d, locked: true, points: 1, schema_version: 3, solution: false, task: false}
#| vscode: {languageId: r}
# Hidden Test Cell
```

## Problem 2

You work at the front desk of a library, and have to manage the printer. Sadly, the printer is old, and regularly jams. You know that your boss will only replace the printer if there is hard evidence, so you decide to record the number of times the printer was able to print successfully **before** it jammed. These values are recorded in `prints` below. Assume that these values are independent and identically distributed.

```{r}
#| deletable: false
#| editable: false
#| nbgrader: {cell_type: code, checksum: 33d0a125ec7a7a3dad34cd46c803bf69, grade: false, grade_id: cell-2b296ffaf48368cd, locked: true, schema_version: 3, solution: false, task: false}
#| vscode: {languageId: r}
prints = c(12, 9, 3, 17, 5, 5, 3, 7, 0, 5, 17, 6, 9, 1, 13)
```

**Part A** Let $X$ be a random variable for the number of times the printer was able to print before it jammed. Then $X$ follows a Geometric distribution with some unknown probability $p$, where $X_i \in {0, 1, 2, \dots}$. We want to estimate $p$, that is the probability that the printer fails. Using the method of moments, what is the estimate $\hat{p}$?

1. $\dfrac{1}{\bar{X}}$
2. $\dfrac{1}{\bar{X}+1}$
3. $\dfrac{\bar{X}}{\bar{X}+1}$
4. $\dfrac{\bar{X}}{1-\bar{X}}$
5. None of the above.

Save the NUMBER corresponding to the correct answer into the variable `p2.a`.

```{r}
#| deletable: false
#| nbgrader: {cell_type: code, checksum: 459914beff70b21439aea04034f5384d, grade: false, grade_id: cell-efc73ab51f772fb8, locked: false, schema_version: 3, solution: true, task: false}
#| vscode: {languageId: r}
p2.a = NA

# your code here
p2.a =2
```

```{r}
#| deletable: false
#| editable: false
#| nbgrader: {cell_type: code, checksum: 2a514f163f318e5e8824754ac0102a46, grade: true, grade_id: cell-877dddf3cb5c2df5, locked: true, points: 1, schema_version: 3, solution: false, task: false}
#| vscode: {languageId: r}
# Hidden Test Cell
if(!test_that("Make sure your answer is the NUMBER of the correct answer", expect_is(p2.a, "numeric"))){
    print("Make sure your answer is the number corresponding to your answer.")
}
```

**Part B** Given your solution to **Part A**, calculate the estimated probability for the recorded data. Save this value into `p2.b`. (Include at least 2 decimal places.)

```{r}
#| deletable: false
#| nbgrader: {cell_type: code, checksum: 20cdb31d2ecac82d6cf17f697c0b4f71, grade: false, grade_id: cell-0219e0276879aa9c, locked: false, schema_version: 3, solution: true, task: false}
#| vscode: {languageId: r}
p2.b = NA

# your code here
phat<-1/ (mean(prints)+1)
p2.b = round(phat,2)
p2.b 
```

```{r}
#| deletable: false
#| editable: false
#| nbgrader: {cell_type: code, checksum: 80404725fb90eb9a7c133dcb26f46cb4, grade: true, grade_id: cell-c7827c9f732779f8, locked: true, points: 1, schema_version: 3, solution: false, task: false}
#| vscode: {languageId: r}
# Hidden Test Cell
```

**Part C** Let's try to visualize how well our estimator actually fits the data. 

To create a histogram of the "prints" data, one only needs to type <code>hist(prints)</code>. In order to store the results in a way the autograder can interpret, we need to store this in a variable like this: <code>my_hist_var=hist(prints)</code>

Recall that the y-axis of a histogram can be scaled in many different ways.
In order to make it so that the total area of each bar is the proportion of observations observed in the corresponding bin, we need to add "prob=T" (probability equals true) to the histogram command. This is called a "density histogram".

Finally, to control the way the bins break on the x-axis, we need to give a sequence of break values. As an example unrelated to this exercise, we might type <code>my_hist_var = hist(mydata,prob=T,breaks=seq(1,8,1))</code> to put breaks at 1,2, ..., 8.

For this exercise,  create a density histogram of the `prints` values and store it in variable `p2.c.hist`. Center the bars on the integers 0 through 17 by making breaks at -0.5, 0.5, 1.5, ..., 17.5. Add the curve of a geometric distribution, with the probability of what you estimated in **Part B**, ontop of your histogram. Think about how well this curve approximates the distribution.

```{r}
#| deletable: false
#| nbgrader: {cell_type: code, checksum: 4268f9975bff034f4cbe5528391da063, grade: false, grade_id: cell-bea143bf92a0563d, locked: false, schema_version: 3, solution: true, task: false}
#| vscode: {languageId: r}
p2.c.hist = NA

# your code here
br = seq(-0.5,17.5,1)
p2.c.hist = hist(prints,probability = T,breaks = br)
x<-seq(0,17,1)
fx<-(1-phat)^x*phat
lines(x,fx,col=4)
```

```{r}
#| deletable: false
#| editable: false
#| nbgrader: {cell_type: code, checksum: 543eb0cd06450deccc9751950291c66e, grade: true, grade_id: cell-a46e96bac036f007, locked: true, points: 1, schema_version: 3, solution: false, task: false}
#| vscode: {languageId: r}
# Hidden Test Cell
```

**Part D** what is the *true* underlying probability of this distribution? Save your answer as `p2.d` and round your answer to two decimal places. If it's not possible to answer this question, save your answer as the string `"not possible"`.

```{r}
#| deletable: false
#| nbgrader: {cell_type: code, checksum: ebc5453322de8a29dc9f4dcb3cee3019, grade: false, grade_id: cell-ecb0de26dd3fad02, locked: false, schema_version: 3, solution: true, task: false}
#| vscode: {languageId: r}
p2.d = NA

# your code here
p2.d = "not possible"
```

```{r}
#| deletable: false
#| editable: false
#| nbgrader: {cell_type: code, checksum: 61359141df8cadc0835370d66b9572d0, grade: true, grade_id: cell-feef83956abbdc3a, locked: true, points: 1, schema_version: 3, solution: false, task: false}
#| vscode: {languageId: r}
# Hidden Test Cell
```

