---
title: "Probability, Expectation, and Variance"
---

[refrigerator]{.column-margin}


::: {#exr-01-refrigerator }
## Expected refrigerator price
A big box store sells three different models of refrigerators having 
20, 21.5, and 26.8 cubic feet of usable storage space, 
respectively. Let X = the amount of storage space purchased by the next 
customer to buy a freezer. Suppose that X has probability mass function given by

| x (cubic feet) | 20  | 21.5 | 26.8 |
|----------------|-----|------|------|
| $\mathbb{P}(x)$| 0.1 | 0.7  | 0.2  |
: PMF for refrigerator storage space {#tbl-refrigerator-pmf }

If the price of a refrigerator with $X$ cubic feet of capacity is 
$PRICE=(20X+100)$, what is the expected price paid by the next customer to buy a freezer?
:::

::: {.callout-tip}
## Solution

1. Solution Strategy: 
    1. we need to find the expected value of PRICE
    2. PRICE is a linear transformation of the X variable
    3. The expectation can then be calculated as usual ie 
       $$
       \sum_i P(x_i) \cdot PRICE(x_i)
       $$ {#eq-refrigerator-expected-price }

```{r}

data <- data.frame(
  x   = c(20,21.5,26.8),
  p_x = c(0.1, 0.7, 0.2)
)
data


Expecred_Price <- function(data) {
  sum((20*data$x+100) * data$p_x) #implementation of @eq-refrigerator-expected-price
}
Expecred_Price(data)
```

:::

[Circuit board]{.column-margin} 

::: {#exr-02-circuit-board }
## Circuit board lifetimes
In proof testing of circuit boards, the probability that any particular diode will fail is 0.03 . Suppose a circuit board contains 150 diodes and that each one performs or fails independently of all others.

[How many diodes would you expect to fail]{.mark}, and what is the [standard deviation]{.mark} of the number that are expected to fail?

:::


::: {.callout-tip}
## Solution

1. Solution Strategy: 
    1. Let $X$ = number of diodes that fail. Since each diode is an iid Bernoulli trials with fixed probability of failure and independence. Their sum will follow a Binomial distribution.
    2. $\therefore X \sim \mathcal{Bin}(n = 150, p = 0.03)$
    3. The expectation and standard deviation of a binomial distribution are given by:
       - $E(X) = n \cdot p \qquad SD(X) = \sqrt{n \cdot p \cdot (1 - p)}$

```{r}
n <- 150
p <- 0.03
expected_failures <- n * p
sd_failures <- sqrt(n * p * (1 - p))
expected_failures
sd_failures
```
:::


---

[Skink dispersal]{.column-margin}


::: {#exr-03-skink-dispersal }
## Skink dispersal distances

Let $X$ denote the distance, in meters, that an animal moves from its birth site to the first territorial vacancy it encounters. Suppose that for Solomon Island prehensile-tailed skinks, $X$ has an exponential distribution with rate parameter  $\lambda = 0.014$.

What is the probability that the distance exceeds the mean distance by more than 2 standard deviations?
:::


::: {.callout-caution}
## Check Parameterization

The best way to mess up is to get the parameterization of the exponential distribution wrong. 
Even worse if we also have a poison in the mix. 
Jem cautions us about this several times.

i.e. do we need to use $\frac{1}{\lambda}$?)

PDF from handout:
$$
f(x) = \lambda e^{-\lambda x} \mathbb{I}_{x \geq 0}(x)
$$

Does R parameterize the same way?
:::

::: {.callout-tip}
## Solution

- strategy:
    1. Given: $X \sim \mathcal{Exp}(rate = 0.014)$  
    2. We have $\mu$ and $\sigma$  from 

:::